---
title: 'Lecture 5: Data Wrangling'
author: "MT"
date: "2025-03-21"
output:
  html_document: default
  md_document:
    variant: gfm
  pdf_document: default
  word_document: default
---

Data wrangling is manipulating data, arranging columns, etc. 
Tidyverse! - 8 packages in one 

```{r}
library(tidyverse)
```

What we will cover:  

  - mutate()
  - select()
  - filter()
  - piping %>%
  - summarise()
  _ group_by()
  - joining
  - pivoting
  - integration with plotting 
  
```{r}
microbiomefungi <- read.csv("Bull_richness.csv")
str(microbiomefungi)
```


#### Selecting
selects certain columns out of a dataset
This isn't that easy to do in base R. 
```{r}
microbiomefungi2 <- select(microbiomefungi, SampleID, Crop, Compartment:Fungicide, richness)
```


#### Filter
essentially the same as subsetting in base R 
```{r}
head(filter(microbiomefungi2, Treatment == "Conv.")) #head func is just to output first 5 rows for demonstration

# can use & 
head(filter(microbiomefungi2, Treatment == "Conv." & Fungicide == "C"))

# can use or |
head(filter(microbiomefungi2, Sample == "A" | Sample == "B")) #samples a or b
```


#### Mutate 
Create a new column 
```{r}
microbiomefungi2$logRich <- log(microbiomefungi2$richness) #add a col in base R

head(mutate(microbiomefungi2, logRich2 = log(richness))) #this does the same thing as adding a col in base R
```


#### Piping
inserts output from one function as input into the next function 
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>%
  filter(Treatment == "Conv.") %>%
  mutate(logRich2 = log(richness)) %>%
  head()         
```

#### Summarise
generate summary stats from our data
```{r, }
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich)) # calculating overall mean log richness within the conventionally managed treatment
```

even more summary stats all at once
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  filter(Treatment == "Conv.") %>% # subsetting to only include the conventional treatment
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```


#### Group_by
calculate mean, std dev for a group rather than filtering 
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group (each unique treatment and fungicide combination)
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n))
```


#### Combine with GGPlot
these functions work well together with ggplot
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  mutate(logRich = log(richness)) %>% # creating a new column of the log richness
  summarise(Mean.rich = mean(logRich), # calculating the mean richness, stdeviation, and standard error
            n = n(), 
            sd.dev = sd(logRich)) %>%
  mutate(std.err = sd.dev/sqrt(n)) %>%
  ggplot(aes(x = Fungicide, y = Mean.rich)) + # adding in a ggplot
  geom_bar(stat="identity") +
  geom_errorbar( aes(x=Fungicide, ymin=Mean.rich-std.err, ymax=Mean.rich+std.err), width=0.4) +
  theme_minimal() +
  xlab("") +
  ylab("Log Richness") +
  facet_wrap(~Treatment)
```


#### Joining
combine multiple datasets using a common variable 

More info can be found [here](https://dplyr.tidyverse.org/reference/mutate-joins.html)

we can easily do this with the joining functions. They include:

- left_join() - Keep all rows of X and add matching rows from Y. Any rows in Y that don’t match X are excluded.
- right_join() - reverse of left_join() 
-inner_join() - only keep rows that are common to both X AND Y, remove everything else. 
- full_join() - Keep any columns that are in either X or Y

First, lets split the data into taxonomy and metadata. 
```{r}
# selecting just the richness and sample ID
richness <- microbiomefungi %>%
  select(SampleID, richness)

# selecting columns that don't include the richness
metadata <- microbiomefungi %>% 
  select(SampleID, Fungicide, Crop, Compartment, GrowthStage, Treatment, Rep, Sample)

head(richness)
head(metadata)
```

left join

- "by" is what the datasets have in common, the col they are joined by
```{r}
head(left_join(metadata, richness, by = "SampleID")) # adding the richness data to the metadata based on on the common column of sampleID
```


#### Pivoting
converting dataset from wide to long view 
Wide data can be useful, but long is usually better for analysis 

More info can be found here: https://tidyr.tidyverse.org/reference/pivot_wider.html

output in long format
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% 
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) # calculates the mean per Treatment and Fungicide 
```

pivot wider - the data is in wide format now
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  filter(Class == "Sordariomycetes") %>%
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) # pivot to wide format
```
names_from: converts the values from that column into new column headers


Can use this to easily take the difference between means of control and fungicide treatments 
```{r}
microbiomefungi %>%
  select(SampleID, Crop, Compartment:Fungicide, richness) %>% # selecting columns  
  group_by(Treatment, Fungicide) %>% # grouping by treatment and fungicide to later calculate summary stats by group
  summarise(Mean = mean(richness)) %>% # calculates the mean per Treatment and Fungicide 
  pivot_wider(names_from = Fungicide, values_from = Mean) %>% # pivot to wide format
  mutate(diff.fungicide = C - F) %>%  # calculate the difference between the means. 
  ggplot(aes(x = Treatment, y = diff.fungicide)) + # Plot it 
  geom_col() +
  theme_minimal() +
  xlab("") +
  ylab("Difference in average species richness")
```


