---
title: "Coding Challenge 4: Markdown"
author: "Madelyn Thompson"
date: "2025-02-27"
output:
  md_document:     
  variant: gfm 
  word_document:
  html_document: 
    toc: TRUE
    toc_float: TRUE
---


### Question 1 

  a. A YAML header is the section of the top of .rmd code that sets the title, output options, and other global settings. You can set the output as a html, word, pdf, or github flavored markdown document. You can also set a table of contents in the YAML header. 
  b. Literate programming is writing code in a way that is human-readable and easy to understand. It is a key part of a reproducible workflow. 
  

### Question 2

Data are published here: 

[Noel, Z.A., Roze, L.V., Breunig, M., Trail, F. 2022. Endophytic fungi aspromising biocontrol agent to protect wheat from Fusarium graminearum head blight. Plant Disease.](https://apsjournals.apsnet.org/doi/10.1094/PDIS-06-21-1253-RE)


Load in the MycotinData.csv dataframe.  
```{r Load in MycotinData.csv dataframe}
# load in data, assigning na with <NA>
mycotoxindf <- read.csv("MycotoxinData.csv", na.strings = "na")
str(mycotoxindf) #check that it's right 
```

Load in packages and colorblind palette. 
```{r cbbpalette}
library(ggplot2)
library(ggpubr)
#color blind palette vector 
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Create DON~Treatment boxplot. 
```{r DON boxplot}
# relevel the df so that NTC is the first group in the plot 
mycotoxindf$Treatment <- factor(mycotoxindf$Treatment, levels = c("NTC", "Fg", "Fg + 37", "Fg + 40", "Fg + 70"))

#create a boxplot of DON~Treatment 
DON.bp <- ggplot(mycotoxindf, aes(x= Treatment, y = DON, fill = Cultivar)) +
  geom_boxplot(outliers = F) + 
  geom_point(shape = 21, alpha = 0.6, position = position_jitter(0.25)) +
  xlab(" ") +
  ylab("DON (ppm)") +
  scale_fill_manual(values = c(cbbPalette[3], cbbPalette[4])) +
  theme_classic() +
  facet_wrap(~Cultivar)
```

Create Mass~Treatment boxplot.
```{r Mass boxplot}
mass.bp <- ggplot(mycotoxindf, aes(x= Treatment, y = MassperSeed_mg , fill = Cultivar)) +
  geom_boxplot(outliers = F) + 
  geom_point(shape = 21, alpha = 0.6, position = position_jitter(0.25)) +
  xlab(" ") +
  ylab("Seed Mass (mg)") +
  scale_fill_manual(values = c(cbbPalette[3], cbbPalette[4])) +
  theme_classic() +
  facet_wrap(~Cultivar)
```

Create X15adon~Treatment boxplot.
```{r x15adon boxplot}
X15ADON.bp <- ggplot(mycotoxindf, aes(x= Treatment, y = X15ADON , fill = Cultivar)) +
  geom_boxplot(outliers = FALSE) + 
  geom_point(shape = 21, alpha = 0.6, position = position_jitter(0.25)) +
  xlab(" ") +
  ylab("15ADON") +
  scale_fill_manual(values = c(cbbPalette[3], cbbPalette[4])) +
  theme_classic() +
  facet_wrap(~Cultivar)
```

Use ggarange to combine all figures and add p-values to figures
```{r combined figure, warning = FALSE, fig.height=10}
# adds t-test pairwise p-values to each of the figures
X15ADON.bp.ttest <- X15ADON.bp +
  geom_pwc(aes(group = Treatment), method = "t.test", label = "p.adj.format") 

mass.bp.ttest <- mass.bp +
  geom_pwc(aes(group = Treatment), method = "t.test", label = "p.adj.format") 

DON.bp.ttest <- DON.bp +
  geom_pwc(aes(group = Treatment), method = "t.test", label = "p.adj.format") 

#combines those three plots with the ttest pvalues into a single plot with 3 panels
figure.ttest <- ggarrange(
  DON.bp.ttest, 
  mass.bp.ttest, 
  X15ADON.bp.ttest,
  nrow = 3, 
  ncol = 1, 
  labels = c("A)", "B)", "C)"), 
  common.legend = TRUE, 
  legend = "bottom"
)
figure.ttest
```





