---
title: "Lecture 6 Iterations and Functions"
author: "Madelyn Thompson"
date: "2025-03-26"
output:
  pdf_document: default
  md_document: 
    variant: gfm
  html_document:
---

load in packages
```{r}
library(ggplot2)
library(tidyverse)
```


## Functions

convert f to c
```{r}
# (5*(degree_f - 32)/9)
(5*(89 - 32)/9) #could lead to copy and paste errors if you need to run this code a lot 

#Solution- make a function! 

F_to_C <- function(f_temp){
  celsius <- (5*(f_temp - 32)/9)
  return(celsius)
}

F_to_C(32)
  
```


## Iterations - Loops! 

can prevent copy and paste errors 

### Interation Functions
```{r}
#repeat elements easily
rep("A", 3) #repeats value A 3 times
rep(c("A", "B"), 10) #repeats 10 times A B A B 
rep(c(1,2,5,3), 4, each = 5) #repeats sequence in order, 5 times for each number

#write sequences of numbers easily
1:7
seq(from = 1, to = 7)
seq(from = 0, to = 10, by = 2) #counts by twos 

#make a sequence of numbers based on a character vector, helpful in loops
LETTERS #built in character vector of each letter in the alphabet
seq_along(LETTERS) #returns 1 - 26 for each letter in the alphabet 
```


### The For Loop !
```{r}
for (i in 1:10){
  print(i*2)
}
  
#can use a function within a loop
for (i in -30:100){
  result <- F_to_C(i)
  print(result)
}

#what if you want to output the loop into an object 
celsius.df <- NULL #create an empty object
for (i in -30:100){
  result <- data.frame(F_to_C(i), i) #create a one row df 
  celsius.df <- rbind.data.frame(celsius.df, result) #add newest row to the celcius df in each loop 
}
```


## Real world For Loop example 

Dose response curve: Nonlinear regression with s-shaped curve
We are trying to find EC50 (where fungicide reduces pop by half)

Important cols: 

- isolate
- concentration of chemical
- relgrowth (col diameter compared to control)

Goal: model each isolate separately


```{r, results = "hide", warning = FALSE, message = FALSE}
library(drc)
# EC50.data <- read.csv("Lectures/Lecture6Iterations&Functions/EC50_all.csv")
EC50.data <- read.csv("EC50_all.csv")

# regression for one isolate
isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == "ILSO_5-41c"] ~ 
        EC50.data$conc[EC50.data$is == "ILSO_5-41c"], 
                       fct = LL.4(fixed = c(NA, NA, NA, NA), 
                                  names = c("Slope", "Lower", "Upper", "EC50")), 
                       na.action = na.omit)
    # outputs the summary of the paramters including the estimate, standard
    # error, t-value, and p-value outputs it into a data frame called
    # summary.mef.fit for 'summary of fit'
    summary.fit <- data.frame(summary(isolate1)[[3]])
    # outputs the summary of just the EC50 data including the estimate, standard
    # error, upper and lower bounds of the 95% confidence intervals around the
    # EC50
    EC50 <- ED(isolate1, respLev = c(50), type = "relative", 
        interval = "delta")[[1]]
    
# turn this into a loop so it does it for each isolate in the dataset and outputs EC50 
nm <- unique(EC50.data$is) #finds unique values of char vector (in this case within "is" col of our df)
nm # outputs all the isolate names

for (i in seq_along(nm)) {
  isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == nm[i]] ~ 
        EC50.data$conc[EC50.data$is == nm[i]], 
                       fct = LL.4(fixed = c(NA, NA, NA, NA), 
                                  names = c("Slope", "Lower", "Upper", "EC50")), 
                       na.action = na.omit)
    # outputs the summary of the paramters including the estimate, standard
    # error, t-value, and p-value outputs it into a data frame called
    # summary.mef.fit for 'summary of fit'
    summary.fit <- data.frame(summary(isolate1)[[3]])
    # outputs the summary of just the EC50 data including the estimate, standard
    # error, upper and lower bounds of the 95% confidence intervals around the
    # EC50
    EC50 <- ED(isolate1, respLev = c(50), type = "relative", 
        interval = "delta")[[1]]
    EC50
}

```

Make this more useful by saving results as an object
```{r, results = "hide", warning = FALSE}
EC50.ll4 <- NULL

for (i in seq_along(nm)) {
  isolate1 <- drm(100 * EC50.data$relgrowth[EC50.data$is == nm[[i]]] ~ 
        EC50.data$conc[EC50.data$is == nm[[i]]], 
                       fct = LL.4(fixed = c(NA, NA, NA, NA), 
                                  names = c("Slope", "Lower", "Upper", "EC50")), 
                       na.action = na.omit)
    # outputs the summary of the paramters including the estimate, standard
    # error, t-value, and p-value outputs it into a data frame called
    # summary.mef.fit for 'summary of fit'
    summary.fit <- data.frame(summary(isolate1)[[3]])
    # outputs the summary of just the EC50 data including the estimate, standard
    # error, upper and lower bounds of the 95% confidence intervals around the
    # EC50
    EC50 <- ED(isolate1, respLev = c(50), type = "relative", 
        interval = "delta")[[1]]
    isolate.ec_i <- data.frame(nm[[i]], EC50) #df with isolate name and EC50 value 
    EC50.ll4 <- rbind.data.frame(EC50.ll4, isolate.ec_i) #Each iteration, the newly generated row is appended to this dataframe 
}
```

plot it
```{r}
ggplot(EC50.ll4, aes(x = EC50)) + geom_histogram() + theme_classic()
```

#### Loops in TidyVerse

Can do this another way by using map function in tidyverse 

- Group by isolate
- nest() - creates a subdataframe for each isolate (the grouped var)
- mutate Creates col called ll.4.mod 
- map() - iterates through data, basically a for loop function
- period '.' inherits all of the data from those subtables
- results from the map loop are outputted into the ll.4.mod col
- mutate again to estimate EC50 from 11.4.mod col and output values into the new col EC50
- unnest it and pull it out of subtables - just unnesting EC50 var 

Fancy way to iterate
```{r, results = "hide", warning = FALSE}
EC50.data %>%
  group_by(is) %>%                                         
  nest() %>%                                
  mutate(ll.4.mod = map(data, ~drm(.$relgrowth ~ .$conc,        
                              fct = LL.4(fixed = c(NA, NA, NA, NA), 
                                         names = c("Slope", "Lower", "Upper", "EC50"))))) %>%
  mutate(ec50 = map(ll.4.mod, ~ED(., 
                              respLev = c(50), 
                              type = "relative",
                              interval = "delta")[[1]])) %>%
  unnest(ec50)
```




