---
title: "Coding Challenge 7 - Linear Models"
author: "Madelyn Thompson"
date: "2025-04-03"
output:
  pdf_document: default
  md_document: 
    variant: gfm
  html_document: default
---

#make pdf not run off the page? 
```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## 1. 4 pts. Read in the data called “PlantEmergence.csv” using a relative file path and load the following libraries. tidyverse, lme4, emmeans, multcomp, and multcompView. Turn the Treatment , DaysAfterPlanting and Rep into factors using the function as.factor
STAND$Treatment <- as.factor(STAND$Treatment) # example shown here.
```{r}
#load in packages
library(tidyverse)
library(lme4)
library(emmeans)
library(multcomp)
library(multcompView)
library(ggplot2)

#load in data
Plantdf <- read.csv("CodingChallenges/PlantEmergence.csv")
Plantdf$Treatment <- as.factor(Plantdf$Treatment)
Plantdf$DaysAfterPlanting <- as.factor(Plantdf$DaysAfterPlanting)
Plantdf$Rep <- as.factor(Plantdf$Rep)
```


## 2. 5 pts. Fit a linear model to predict Emergence using Treatment and
DaysAfterPlanting along with the interaction. Provide the summary of the linear
model and ANOVA results.
```{r}
#visualize the data
ggplot(data= Plantdf, aes(x= Treatment, y= Emergence, fill = DaysAfterPlanting)) +
  geom_boxplot()

#treatment within Days after planting interaction
lm <- lm(Emergence~Treatment+DaysAfterPlanting+Treatment:DaysAfterPlanting, data =Plantdf)
summary(lm)
anova(lm)

#is the interaction different if it's written the other way? 
lm2 <- lm(Emergence~Treatment+DaysAfterPlanting+DaysAfterPlanting:Treatment, data =Plantdf)
summary(lm2)
anova(lm2)
#no these are the same results
```


## 3. 5 pts. Based on the results of the linear model in question 2, do you need to fit the
interaction term? Provide a simplified linear model without the interaction term but still testing both main effects. Provide the summary and ANOVA results. Then, interpret the intercept and the coefficient for Treatment 2.

**The treatment x days after planting interaction was not significant. ** 
```{r}
lm3 <- lm(Emergence~Treatment+DaysAfterPlanting, data =Plantdf)
summary(lm3)
anova(lm3)
```

**The intercept is the least squared mean emergence of treatment one at 7 days (182.163). The coeffecient is the difference in treatment 2 mean emergence from the reference.** 



## 4. 5 pts. Calculate the least square means for Treatment using the emmeans package and
perform a Tukey separation with the compact letter display using the cld function.
Interpret the results.

```{r}
lsmeans <- emmeans(lm3, ~Treatment)   #runs least squared means (means estimated by the linear model)

results_lsmeans <- cld(lsmeans, alpha = 0.05, details = TRUE) # outputs which groups are different from each other 
# .group number shows which ones are the same and sig different from each other 
results_lsmeans
```

**Treatment 2 was significantly lower than all other groups with a mean of 55.6 plants emerged. Treatments 6, 3, and 5 had the highest means and were significantly higher than treatments 7 and 2. Treatments 1, 9, 4, 8 were only significantly different from treatment 1**



## 5. 4 pts. The provided function lets you dynamically add a linear model plus one factor from that model and plots a bar chart with letters denoting treatment differences. Use this model to generate the plot shown below. Explain the significance of the letters.

```{r}
plot_cldbars_onefactor <- function(lm_model, factor) {
  data <- lm_model$model
  variables <- colnames(lm_model$model)
  dependent_var <- variables[1]
  independent_var <- variables[2:length(variables)]
  lsmeans <- emmeans(lm_model, as.formula(paste("~", factor))) # estimate
  lsmeans
  Results_lsmeans <- cld(lsmeans, alpha = 0.05, reversed = TRUE, details =
  TRUE, Letters = letters) # contrast with Tukey adjustment by default.
  # Extracting the letters for the bars
  sig.diff.letters <- data.frame(Results_lsmeans$emmeans[,1],
  str_trim(Results_lsmeans$emmeans[,7]))
  colnames(sig.diff.letters) <- c(factor, "Letters")
  # for plotting with letters from significance test
  ave_stand2 <- lm_model$model %>%
  group_by(!!sym(factor)) %>%
  dplyr::summarize(
  ave.emerge = mean(.data[[dependent_var]], na.rm = TRUE),
  se = sd(.data[[dependent_var]]) / sqrt(n())
  ) %>%
  left_join(sig.diff.letters, by = factor) %>%
  mutate(letter_position = ave.emerge + 10 * se)
  plot <- ggplot(data, aes(x = !! sym(factor), y = !! sym(dependent_var))) +
  stat_summary(fun = mean, geom = "bar") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  ylab("Number of emerged plants") +
  geom_jitter(width = 0.02, alpha = 0.5) +
  geom_text(data = ave_stand2, aes(label = Letters, y = letter_position),
  size = 5) +
  xlab(as.character(factor)) +
  theme_classic()
  return(plot)
}

plot_cldbars_onefactor(lm3, "Treatment")
```

**The letters show which treatments are significantly different from each other. Treatment 2 has a unique letter, c, which means it is significantly different from all other groups. Treatments 3, 5, and 6, are significantly higher than treatments listed with only b and c, but not different from each other or those also listed with a (such as 1, 4,and 8.**


## 6. 2 pts. Generate the gfm .md file along with a .html, .docx, or .pdf. Commit, and push the .md file to github and turn in the .html, .docx, or .pdf to Canvas. Provide me a link here to your GitHub. 

[Link to GitHub](https://github.com/mgt0021/ENTM6820.git)