---
title: 'Coding Challenge 5: Data manipulation'
author: "Madelyn Thompson"
date: "2025-03-21"
output: 
  word_document:
  md_document: 
    variant: gfm
  html_document:
---

```{r}
library(tidyverse)
```


#### Q1. Load in data using relative file path
```{r}
diversity <- read.csv("DiversityData.csv")
str(diversity)

metadata <- read.csv("Metadata.csv")
str(metadata)
```
Is something broken? Why does the chunk run with one file path, but the knitting requires a different file path?  

#### Q2. Join dataframes together by code column
```{r}
alpha <- left_join(metadata, diversity, by = "Code")
head(alpha)
```

#### Q3. Calculate Pielou's Evenness (Shannon / log(richness))
```{r}
alpha_even <- mutate(alpha, evenness = shannon / log(richness))
head(alpha_even)
```

#### Q4. Summarize to find the mean and std error of evenness grouped by crop over time 
```{r}
alpha_average <- alpha_even %>%
  group_by(Crop, Time_Point) %>% 
  summarise(Mean = mean(evenness), n = n(), std.dev = sd(evenness)) %>% 
  mutate(std.err = std.dev/sqrt(n))
alpha_average
```

#### Q5. Calculate the difference between the mean evenness of the crop types and soil at each time point
```{r}
alpha_average2 <- alpha_average %>%
  select(Crop, Time_Point, Mean) %>%
  pivot_wider(names_from = Crop, values_from = Mean) %>% 
  mutate(diff.soybean.even = Soil-Soybean) %>%
  mutate(diff.cotton.even = Soil-Cotton)
alpha_average2
```

#### Q6. Create a plot 
```{r}
alpha_average2 %>%
  select(diff.soybean.even, diff.cotton.even, Time_Point) %>%
  pivot_longer(cols = c(diff.soybean.even, diff.cotton.even), names_to = "diff") %>%
  ggplot(aes(x=Time_Point, y = value, color = diff)) +
  geom_line() +
  xlab("Time(hrs)") +
  ylab("Difference from soil in Pielou's evenness") +
  theme_classic() +
  scale_color_discrete(name = "Crop", labels = c("Cotton", "Soybean"))
```

